Using Scipion for Stream Image Processing at EM Facilities

### 2018-03-21 1.introduction.tex
w

3D electron microscopy (3DEM)  can provide rich information about structural characteristics of macromolecular complexes. The field is under a major transformation due to the arrival of better microscopes, new large area detectors and automatization \citep{kuhlbrandt2014a, Kuhlbrandt2014b}. These improvements make possible that a single microscope generates high quality data sets on the order of terabytes per day \citep{Saibil2015} while working for several days without interruption. Major challenges faced in the field are: (1) an efficient management of these huge datasets and their corresponding image processing workflows; (2) a raising interest from scientists outside the field who might lack the skills of a experienced microscopist and; (3) the need for an effective data and metadata transfer in an increasingly distributed and collaborative environment. This situation strongly demands the application of techniques that are highly automated to allow high throughput.

@@

Clearly, the future of projects that require high-end microscopes is through centralized microscopy facilities, in which automatic acquired data needs to be monitored by users and facility staff. [**JMRT VERY LARGE AND TWISTED SENTENCE \scipion has been redesigned with this scenario in mind in which the early detection of issues may save a data acquisition shift which otherwise needs to be repeated after a new application that may be granted many months later]. In this way, \scipion starts the processing of the movies as they are being acquired with the double aim of checking for possible data collection errors and execute a first user-tailored image processing workflow that provides users with an accurate measurement of the acquired data quality.

Many software packages have been developed to provide automated data collection, such as SerialEM \citep{Mastronarde2005}, Leginon \citep{Suloway2009}, UCSF-Tomography \citep{Zheng2007}, Tom Toolbox \citep{Nickell2005}, EPU \citep{EPU}, Latitude in Digital Micrograph \citep{Latitude}, or EMMenu \citep{emmenu}, etc. However, the next step,  which is to provide image processing while  data are being collected, is not so well addressed. In fact, in most places this is accomplished through home made scripts (i.e. \citep{Pichkur2018}). Exceptions are UCSFImage4 \citep{Li2015} and Focus \citep{Biyani2017}. As we will show in this article, \scipion differs from these solutions in a larger flexibility both from the point of view of users and facility staff.

>>>>

3D electron microscopy (3DEM)  can provide rich information about structural characteristics of macromolecular complexes. The field is under a major transformation due to the arrival of better microscopes, new large area detectors and automatization \citep{kuhlbrandt2014a, Kuhlbrandt2014b}. These improvements make possible that a single microscope generates high quality data sets on the order of terabytes per day \citep{Saibil2015} while working for several days without interruption. Major challenges faced in the field are: (1) an efficient management of these huge datasets and their corresponding image processing workflows; (2) a raising interest from scientists outside the field who might lack the skills of a experienced microscopist and; (3) the need for an effective data and metadata transfer in an increasingly distributed and collaborative environment. %This situation strongly demands the application of techniques that are highly automated to allow high throughput.

@@

Clearly, the future of projects that require high-end microscopes is through centralized microscopy facilities, in which automatic acquired data needs to be monitored by users and facility staff.  \scipion has been redesigned with this scenario in mind. % in which the early detection of issues may save a data acquisition shift which otherwise needs to be repeated after a new application that may be granted many months later. 
In this way, \scipion starts the processing of the movies as they are being acquired with the double aim of checking for possible data collection errors and execute a first user-tailored image processing workflow that provides users with an accurate measurement of the acquired data quality.

Many software packages have been developed to provide automated data collection, such as SerialEM \citep{Mastronarde2005}, Leginon \citep{Suloway2009}, UCSF-Tomography \citep{Zheng2007}, Tom Toolbox \citep{Nickell2005}, EPU \citep{EPU}, Latitude in Digital Micrograph \citep{Latitude}, EMMenu \citep{emmenu}, etc. However, the next step,  which is to provide image processing while  data are being collected, is not so well addressed. In fact, in most places this is accomplished through home made scripts \citep[i.e.][]{Pichkur2018}. %(i.e. \citet{Pichkur2018}).
Exceptions are UCSFImage4 \citep{Li2015} and Focus \citep{Biyani2017}. As we will show in this article, \scipion differs from these solutions in a larger flexibility both from the point of view of users and facility staff.

### 2018-03-21 2.overall.tex
w

\section{Main Modifications introduced in  \scipion: Streaming, Monitors and Consensus}
\label{overall}
[**ROB this section does not flow smoothly, needs a second pass]

In the following we describe the main modifications made to \scipion in order to facilitate its deployment and execution in centralized facilities.

The main change introduced in \scipion is \emph{stream processing}. Stream processing has been designed to analyze and react on real-time so that enables analysis of data as it is produced.
In this way, a new level of parallelization has been reached since the  data generated by the microscope continuously flows through a chain of protocols. For example, as movies are imported to the system they are aligned and their CTF is estimated. In \scipion version 1.1, users may find  the first steps in the image processing adapted to streaming, including movie alignment and CTF estimation. We have also implemented particle picking and extraction, that are going to be released soon in the new version 1.2. In the developers version, several 2D classification algorithms are being adapted for this new way of processing.

\scipion stream implementation is based on periodic updates of the protocol input and output data during the image processing. 
Each protocol run in Scipion is an independent process that can additionally create more processing threads. Each run that works in streaming mode has one of it threads monitoring input and updating output, that could overlap with the computing. The separation between the execution engine and the processing logic of each algorithm marks an important difference between running a script or using \scipion.The latter is a more general and extensible approach that allows users to modify the workflow on the fly without modifying a single code line. Moreover, it also allows to quickly extend the current functionality with new methods.

@@

Since \scipion original publication we have developed many \emph{metaprotocols}, that is, 
protocols that either check the progress of other protocols or compare the results of equivalent protocols.

The first type of metaprotocols are called \emph{monitors} and are used to produce live analysis plots, generate reports or raise alerts when some problems are detected. A monitor example is the \protocol{CTF-monitor}, that checks the computed defocus values for each micrograph as they are generated. CTF-monitor may raise an alert if the defocus values are above or below certain thresholds and continuously generate HTML files so that users and staff may easily follow the data acquisition and processing in-house or remotely. An example of this reports may be seen at \url{http://nolan.cnb.csic.es/scipionbox/} [**ROB NOTE: select a nice one an complete URL]%At the project's end, a review of the processing history is produced. [** URL to report example?]
%This feature was very important since we knew that difference facilities will have different needs and criteria to be fulfilled.

The second set of metaprotocols are grouped under the name \emph{consensus} [**JMRT Consesus protocol existed in the first release, **ROB Ok this is true I need to think about how to rewrite this]. For a given logical step (for example, particle picking) these protocols check if the datasets obtained from the same input data using a different algorithm are consistent. Continuing with the particle picking example, \emph{consensus picking} will compare the particles selected for each one of the executed particle-picking algorithms and produce as output a particle set containing those particles selected by most of the algorithms.

Despite the modifications introduced to \scipion data and process model, the framework still provides full data provenance. Data provenance is an important form of metadata that describes how a particular data set was generated by detailing: the steps, input data and parameters used in the computation. Data provenance guaranties reproducibility and traceability. %Since provenance is solved by Scipion and not Scipionbox we will not discuss here how it is been achieved. We just note that, before choosing which data has to be stored, it is necessary to define how these data have to be structured so that they can be later recovered and understood . %This abstract modeling of the data also simplifies the software interoperability offered by Scipion.

>>>>

\section{Main Modifications introduced in  \scipion: Streaming and Metaprotocols}
\label{overall}

In the following we describe the most important modifications made to \scipion in order to facilitate its deployment and execution in centralized facilities.

\subsection{Stream Processing}
The main change introduced in \scipion is \emph{stream processing}, that is, computing on data directly as it is produced or received.
In this way, a new level of parallelization has been reached since the  data generated by the microscope continuously flows through a chain of protocols. For example, as movies are imported to the system they are aligned and their CTF is estimated. In \scipion version 1.1, users may find  the first steps in the image processing adapted to streaming, including movie alignment and CTF estimation. We have also implemented particle picking and extraction, that are going to be released soon in the new version 1.2. In the developers version, several 2D classification algorithms are being adapted for this new way of processing.

%\scipion is a multilayered application, the first layer handles the interaction with the user (\emph{presentation layer}), the second saves and retrieves the data used by the application  (\emph{data layer}) and finally, the third layer (\emph{business layer}) process the data following the user instructions.  
\scipion stream implementation make extensive use of threads. %A thread is the smallest unit of processing that can be scheduled by an operating system. 
Using threads, a programmer can effectively create two or more tasks that run at the same time and share the protocol resources (i.e. memory). 
%is based on periodic updates of the protocol input and output data during the image processing. 
%Each protocol execution in \scipion is an independent process that can create processing threads. 
Each protocol that works in streaming mode has one of its threads monitoring input and updating output  while other threads take care of the data processing. This clear separation between data access and data processing marks an important difference between running a script or using \scipion. Since the later facilitates a form of computing in which several computations are executed  concurrently, instead of sequentially. Moreover, \scipion workflow engine, allows users to modify the workflow on the fly without modifying a single code line.
%In practice, the latter follows  a more general and extensible approach that allows users to modify the workflow on the fly without modifying a single code line. %Moreover, it also allows to quickly develop new workflows .

%Since allow to a form of computing in which several computations are executed during overlapping time periods—concurrently—instead of sequentially

@@

\subsection{Metaprotocols}
Since \scipion original publication we have developed many \emph{metaprotocols}, that is, 
protocols that either check the progress of other protocols or compare the results of equivalent protocols.

The first type of metaprotocols are called \emph{monitors} and are used to produce live analysis plots, generate reports or raise alerts when some problems are detected. A monitor example is the \protocol{CTF-monitor}, that checks the computed defocus values for each micrograph as they are generated. CTF-monitor may raise an alert if the defocus values are above or below certain thresholds and continuously generate HTML files so that users and staff may easily follow the data acquisition and processing in-house or remotely. An example of this reports may be seen at \url{http://nolan.cnb.csic.es/scipionbox/} [**ROB NOTE: select a nice one an complete URL, I need to ask to javi chichon]%At the project's end, a review of the processing history is produced. [** URL to report example?]
%This feature was very important since we knew that difference facilities will have different needs and criteria to be fulfilled.

The second set of metaprotocols are grouped under the name \emph{consensus}. For a given logical step (for example, particle picking) these protocols check if the datasets obtained from the same input data using a different algorithm are consistent. Continuing with the particle picking example, \emph{consensus picking} will compare the particles selected for each one of the executed particle-picking algorithms and produce as output a particle set containing those particles selected by most of the algorithms.

We would like to finish this section assuring that despite the modifications introduced to \scipion data and process model, the framework provides full data provenance. That is, for any object created by a  protocol, \scipion always annotate how this object has been generated by detailing: the steps, input data and parameters used in the computation. Data provenance guaranties reproducibility and traceability. %Since provenance is solved by Scipion and not Scipionbox we will not discuss here how it is been achieved. We just note that, before choosing which data has to be stored, it is necessary to define how these data have to be structured so that they can be later recovered and understood . %This abstract modeling of the data also simplifies the software interoperability offered by Scipion.

### 2018-03-21 3.workflow.tex
w

So far, \scipion is being used routinely in four Cryo-EM facilities: \emph{Science for life Laboratory} (SciLifeLab, \url{http://www.scilifelab.se}),  \emph{European Synchrotron Radiation Facility} (ESRF, \url{http://www.esrf.eu/}), \emph{Diamond Light Source Synchrotron} (Diamond, \url{http://www.diamond.ac.uk}) and \emph{National Center for Biotechnology} (CNB, \url{http://www.cnb.csic.es}). \scipion original design satisfied the requirements of the CNB, in which \scipion automatically fetches newly recorded movies from a network mounted disk. After this first installation, \scipion was deployed on SciLifeLab (a centralized facility located at Stockholm) where a rigorous booking system is followed. There, \scipion was adapted to interface with this booking platform in order to produce better reports. Lately,  \scipion has been installed in  large synchrotrons  such as ESRF and Diamond  where projects are handled by ISPyB (a customized laboratory management information system -LIMBS- \citep{Delageniere2011}) and data are saved on a distributed file system that heavily penalizes access to lists of files. In this environment, \scipion needs to constantly interchange information with ISPyB, and it cannot longer check if there is a new file in a particular storage device but must wait to be informed by the system that a new file has been created. All these particularities, that will be described in detail in the following, could only be handled thanks to a careful design that prioritizes flexibility at several levels.

>>>>

So far, \scipion is being used routinely in four Cryo-EM facilities: \emph{Science for life Laboratory} (SciLifeLab, \url{http://www.scilifelab.se}),  \emph{European Synchrotron Radiation Facility} (ESRF, \url{http://www.esrf.eu/}), \emph{eBIC facility at Diamond Light Source} (Diamond, \url{http://www.diamond.ac.uk}) and \emph{National Center for Biotechnology} (CNB, \url{http://www.cnb.csic.es}).[**update] \scipion original design satisfied the requirements of the CNB, in which \scipion automatically fetches newly recorded movies from a network mounted disk. After this first installation, \scipion was deployed on SciLifeLab (a centralized facility located at Stockholm) where a rigorous booking system is followed. There, \scipion was adapted to interface with this booking platform in order to produce better reports. Lately,  \scipion has been installed in  large synchrotrons  such as ESRF and Diamond  where projects are handled by ISPyB (a customized laboratory management information system -LIMBS- \citep{Delageniere2011}) and data are saved on a distributed file system that heavily penalizes access to lists of files. In this environment, \scipion needs to constantly interchange information with ISPyB, and it cannot longer check if there is a new file in a particular storage device but must wait to be informed by the system that a new file has been created. All these particularities, that will be described in detail in the following, could only be handled thanks to a careful design that prioritizes flexibility at several levels.

### 2018-03-21 3.workflow_cnb.tex
w

\subsection{CNB}

>>>>

\subsection{CNB}
The National Center for Biotechnology (CNB) forms part of the Spanish National Research Council(CSIC), the largest public research institution in Spain. 
The CNB Cryo-Electron Microscopy Facility is a joint effort of the Centro Nacional de Biotecnología (CNB-CSIC) and the Centro de Investigaciones Biológicas (CIB-CSIC). The mission of the facility is to provide researchers access to the latest cryo-EM technology for high resolution imaging.

### 2018-03-21 3.workflow_diamond.tex
w

\subsection{Diamond Light Source Facility??? or eBIC???}

\begin{verbatim}

We need to cover here differences with CNB setup

 DEVELOPER RELATED
 * Describe infrastructure in diamond: file system (distributed, user permissions),
                                       no graphical output, queues, 
 * Describe Scipion modification:      import (detect new files by socket message)
                                       monitor, report to ISPyB 
                                       scipion headless
                                       execution as script

USER RELATED
 * how users interact with scipion?
 * how data is visualized?
\end{verbatim}

>>>>

\subsection{The eBIC facility at Diamond Light Source}

Diamond Light Source is the UK National Synchrotron User Facility.  As such, it provides access to large-scale experimental facilities to the UK and worldwide scientific communities to conduct experiments that are not possible in their home laboratories.  As a consequence of this national role, Diamond has significant experience at managing many visiting scientists and user experiments, and the associated data and post-processing.  %In the area of Macromolecular Crystallography, this has become a highly automated process which allows users to visualise electron density maps within minutes of the experiment being completed (REF. Aller P et al Methods Mol Biol 2015)
%The recent ``Revolution Resolution'' (REF. Kulbrandt W, Science 2014) in cryoEM has transformed the impact and demand of cryoEM in Structural Biology. However, the instrumentation and running costs of the latest generation of EM necessary for high-end cryoEM is beyond scope of all but the best-funded home laboratories. Due to the huge demand for high-end microscope access, 
The electron Bio-Imaging Centre (eBIC) was set up at Diamond~\citep{diamond2017}. The centre houses four Thermofisher Titan Krios, a Talos Arctica and a Scios dual-beam. All of the transmission electron microscopes are equipped with counting-mode direct electron detectors and Volta phase plates.

Diamond has thousands of users every year, both academic and commercial, and it is vital that their data is kept secure and confidential. A user management system has been developed which allocates each user with a unique federal identity (IDs), and associates those user IDs with the experiments in which they partake in. Visits are structured such that all raw and processed data collected during an experiment are kept in one directory, and only the associated visitors have read and write permissions in that directory. Users may access their data through ISPyB \citep{ispb2011},  a Laboratory Information Management System (LIMS) combining sample tracking and experiment reporting  %The link between the user and their visits enables data management such as archiving data to a long-term tape storage system, or the Laboratory Information Management System ISPyB (REF. Delageniere et al Bioinformatics 2011). In contrast to the CNB setup, Diamond has integrated the microscopes in much the same fashion to that of the beamlines at the facility, as is shown in figure X

Data is collected onto a local machine %in a similar way to the CNB setup, but 
and then immediately moved via 10Gbps Ethernet to the associated visit directory in the centralised storage, a multi Petabyte GPFS High Performance Parallel file system.  Once the raw data is present on the GPFS system, high speed interconnects to the central cluster enable rapid data processing. %Services such as these are shared amongst the whole facility allowing for flexibility in their deployment and generally allowing better service.

Once the data has reached the central file system, it is available to be processed by Scipion.  This process is facilitated by the use of a Scipion-workflow-template which enable novice users to trigger semi-customised on the fly processing. This template consists of a small UI (figure XX) which allows the user to put in the specific values pertinent to their experiment, namely the session ID (or visit number), which microscope they are using, and several other key data collection parameters.  Once set this makes use of a global template for Scipion as set up by the eBIC team, and substitutes the appropriate information into it, and then saves it to disk.

Following this Diamond makes use of a central ActiveMQ [** what is this? a queue system] service which is triggered upon the template file being available in the visit directory, and runs a headless (i.e. without UI) version of Scipion which takes the template and executes in streaming mode.  THIS SHOULD PROBABLY INCLUDE SOMETHING ABOUT THE DETECTION OF NEW FILES BY SOCKET MESSAGE, BUT I DON’T KNOW HOW THIS WORKS. Me neither, I thought it polled the filesystem?

To monitor the progress of the automatic processes, the Scipion pipeline at Diamond also includes custom nodes which allow it to interface with ISPyB.  This provides the users with a web interface showing general metadata about their experiment as well as new real-time updates on steps of the process and relevant output such as motion corrected images, drift values and CTF images (Figure XXX)

### 2018-03-21 scipionbox.tex
W
w

%\usepackage{lineno}
\usepackage[update,prepend]{epstopdf}
%\linenumbers
\definecolor{lightgreen}{rgb}{0.75,0.75,0.0}
%\usepackage{caption}
%\usepackage{endfloat}
%\DeclareCaptionFormat{empty}{}
%\captionsetup{format=empty}
\pagestyle{fancy}
\fancyhf{} % clear all header and footer fields
\fancyfoot[R]{\footnotesize  \thepage}
\renewcommand{\headrulewidth}{0pt}
\renewcommand{\footrulewidth}{0pt}
\def\scipionbox{\textit{ScipionBox}\xspace}
\def\scipion{\textit{Scipion}\xspace}
\newcommand{\protocol}[1]{\texttt{#1}}

@@

\begin{document}
\title{ScipionBox: automatic processing for }
\author{XXXXX XXXXX$^{(a)}$, 
        YYYYY YYYYY$^{(b)}$, \\
        zzzzzz$^{(m)}$,
        yyyyyyy$^{(b)}$
        }

@@

%\lyxaddress{{\small \vspace{-0.5cm}$^{(a)}$Escuela Polit\'{e}cnica Superior, Universidad Aut\'{o}noma de Madrid, 28049 Cantoblanco, Madrid, Spain.}}
%\lyxaddress{{\small$^{(g)}$ Biocomputing Unit, National Center for Biotechnology (CSIC), C/ Darwin, 3, Campus Universidad Aut\'{o}noma, 28049 Cantoblanco,
%Madrid, Spain.}}

@@

\newpage
\input{abstract.tex}
\input{introduction.tex}
\input{overall.tex}
\input{workflow.tex}
\input{results.tex}
\input{conclusions.tex}
\input{acknowledgments.tex}
\appendix

>>>>

%\usepackage{lineno}
\usepackage[update,prepend]{epstopdf}
%\linenumbers
\definecolor{lightgreen}{rgb}{0.75,0.75,0.0}
%\usepackage{caption}
%\usepackage{endfloat}
%\DeclareCaptionFormat{empty}{}
%\captionsetup{format=empty}
\pagestyle{fancy}
\fancyhf{} % clear all header and footer fields
\fancyfoot[R]{\footnotesize  \thepage}
\renewcommand{\headrulewidth}{0pt}
\renewcommand{\footrulewidth}{0pt}
\def\scipionbox{\textit{ScipionBox}\xspace}
\def\scipion{\textit{Scipion}\xspace}
\def\emadmin{\textit{EMAdmin}\xspace}
\newcommand{\protocol}[1]{\texttt{#1}}

@@

\begin{document}
\title{Using Scipion for Stream Image Processing at EM Facilities}
\author{XXXX$^{(a)}$,\\
        YYYY$^{(b)}$,\\
        }

@@

\lyxaddress{{\small \vspace{-0.5cm}$^{(a)}$Escuela Polit\'{e}cnica Superior, Universidad Aut\'{o}noma de Madrid, 28049 Cantoblanco, Madrid, Spain.}}
\lyxaddress{{\small$^{(b)}$ Biocomputing Unit, National Center for Biotechnology (CSIC), C/ Darwin, 3, Campus Universidad Aut\'{o}noma, 28049 Cantoblanco, Madrid, Spain.}}

@@

\newpage
\input{0.abstract.tex}
\input{1.introduction.tex}
\input{2.overall.tex}
\input{3.workflow.tex}
\input{4.results.tex}
\input{5.conclusions.tex}
\input{6.acknowledgments.tex}
\appendix
